Wide binaries in Gaia DR1

### 2016-09-25 paper/ms.tex

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized
likelihood (FML) of the stars having the same tangential velocity (numerator)
compared to the FML of the stars having different tangential velocities:
\begin{equation}
  \frac{p(\propm_i, \propm_j)_{\vperp_i = \vperp_j}}
  {p(\propm_i, \propm_j)_{\vperp_i \neq \vperp_j}}.
\end{equation}
When this ratio is large, the pair is a candidate binary star. These likelihood
expressions are marginalized over true tangential velocity and distance for each
star in the pair. For the numerator,
\begin{equation}
  p(\propm_i, \propm_j)_{\vperp_i = \vperp_j} =
    \int \, \dd \vperp \, \dd d_i \, \dd d_j \,
    p(\propm_i, \varpi_i, \vperp, d_i, \matrx{C}_i) \,
    p(\propm_j, \varpi_j, \vperp, d_j, \matrx{C}_j)
\end{equation}
where
\begin{align}
  p(\propm, \varpi, \vperp, d, \matrx{C}) &=
    p(\propm, \varpi \given \vperp, d, \matrx{C}) \, p(\vperp) \, p(d) \\
  p(\propm, \varpi \given \vperp, d, \matrx{C}) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\varpi, \mu_\alpha, \mu_\delta\right)} \\
  \bs{x}_\theta &= \transp{\left(\frac{1}{d}, \frac{v^\perp_\alpha}{d}, \frac{v^\perp_\delta}{d}\right)}
\end{align}
and we assume Gaussian priors on the tangential velocity components, $\vperp=(v^\perp_\alpha,v^\perp_\delta)$, and take the ``Milky Way distance prior'' from Coryn Bailer-Jones
\begin{align}
  p(v^\perp_\alpha,v^\perp_\alpha) = \mathcal{N}(0,25~\kms) \, \mathcal{N}(0,25~\kms)
\end{align}

For the denominator of the likelihood ratio,
\begin{equation}
  p(\propm_i, \propm_j)_{\vperp_i \neq \vperp_j} =
    \int \, \dd \vperp_i \, \dd \vperp_j \, \dd d_i \, \dd d_j \,
    p(\propm_i, \varpi_i, \vperp_i, d_i, \matrx{C}_i) \,
    p(\propm_j, \varpi_j, \vperp_j, d_j, \matrx{C}_j)
\end{equation}

>>>>

% define macros for text
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\rave}{\project{\acronym{RAVE}}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}
\newcommand{\documentname}{\textsl{Article}}
\newcommand{\sectionname}{Section}
\newcommand{\figname}{Figure}

@@

For a pair of stars $(i,j)$, we want to ratio the fully marginalized likelihood
(FML) of the stars having the same 3-space velocity (hypothesis 1,
$\mathcal{L}_1$) over the FML of the stars having different 3-space velocities
(hypothesis 2, $\mathcal{L}_2$). When this ratio is large, the pair is a
candidate binary star. For \gaia\ DR1, we only observe parallax, $\varpi$, and
two proper motion components, $\propm = (\mu_\alpha, \mu_\delta)$ for each star.
Where there is overlap with \rave\ or \apogee, some number of observed radial
velocities, $v_r$, with observed variances $\sigma^2_{v_r}$ (independent of the
\gaia\ data) may also be available. We assume that the uncertainties in these
observed quantites are known with covariance matrix $\matrx{C}$. The likelihoods
$\mathcal{L}_1, \mathcal{L}_2$ are marginalized over true 3-space velocity and
distance for each star in the pair.
\begin{align}
  \mathcal{L}_1 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v} \,
    L_i(\bs{v}, d_i) \,
    L_j(\bs{v}, d_j) \,
    p(\bs{v}) \, p(d_i) \, p(d_j) \\
  \mathcal{L}_2 &=
    \int \, \dd d_i \, \dd d_j \, \dd^3 \bs{v}_i \, \dd^3 \bs{v}_j \,
    L_i(\bs{v}_i, d_i) \,
    L_j(\bs{v}_j, d_j) \,
    p(\bs{v}_i) \, p(\bs{v}_j) \, p(d_i) \, p(d_j) \\
\end{align}
where
\begin{align}
  L(\bs{v}, d) &=
    \left[\det\left(\frac{\matrx{C}^{-1}}{2\pi}\right)\right]^{1/2} \,
    \exp \left[ -\frac{1}{2} \transp{\left(\bs{x} - \bs{x}_\theta \right)} \,
    \matrx{C}^{-1} \,
    \left(\bs{x} - \bs{x}_\theta \right) \right] \\
  \bs{x} &= \transp{\left(\varpi, \mu_\alpha, \mu_\delta, v_r \right)} \\
  \bs{x}_\theta &= \transp{\left(\frac{1}{d}, \frac{v_\alpha}{d},
    \frac{v_\delta}{d}, \tilde{v}_r \right)}
\end{align}
We assume Gaussian priors on the true velocity components and use a \todo{which
one will we use?} distance prior \citep{Astraatmadja:2016} on heliocentric
distance
\begin{align}
  p(\bs{v}) &= \mathcal{N}(0,25)~\kms \\
  p(d) &= \frac{r^2}{2\,L^3} \, \exp\left(-r/L\right)
\end{align}

For hypothesis 1, the integral over velocity is non-trivial because the same
``true'' 3-space velocity vector projects into tangential velocity components
differently depending on sky position when the sky separation is appreciable
($v_\alpha = v_\alpha(\bs{v}, \alpha, \delta)$). For hypothesis 2, the integral
can be split into two the product of two simpler integrals $\mathcal{L}_2 = Q_i
\, Q_j$ where
\begin{equation}
  Q = \int \, \dd d \, \dd^3 \bs{v} \, L(\bs{v}, d) \, p(\bs{v}) \, p(d)
\end{equation}
